---
title: "Fun with distributions"
author: "Justin Chartron"
date: "5/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulating a poisson

R has functions for most commonly used distributions in base.

```{r}
library(ggplot2) # needed for qplot
poisson.values = rpois(200, 25)
poisson.values
qplot(poisson.values)
```

Let's inpt the log-likelihood function and create a test sequence

```{r}
ll = function(x, lambda) {-length(x)*lambda - sum(log(factorial(x))) + log(lambda)*sum(x) }
testlambda <- seq(1, 50, 0.1)
```

We can now use a functional to calculated a log likelihood for every test value of lambda!

```{r}
testll <- sapply(testlambda, function(y) ll(poisson.values,y))
plot(testlambda, testll)
```
We can see from these graphs that the likelihood function has a single maximum, so solving for the derivative of the log-likelihood function will provide a maximum.

The MASS library provides many maximum likelihood estimators for common distributions
```{r}
library(MASS)
fitdistr(poisson.values, "poisson")
```

For completeness, we'd expect the maximum likelihood estimator to match mean mean. Let's test.

```{r}
mean(poisson.values)
```
Pretty anticlimactic for the Poisson distribution, but this is how you can derive a maximum likelihood estimator for other distribution paramaters.

Now we can simulate what we would expect to see in a sequencing experiment. Let's suppose we have 3 biological replicates (many RNA-seq experiments only have 2). Suppose the true poisson parameter is 42. Can we estimate this using only 3 values?

```{r}
trial = vector(mode = "list", length = 5000) # Create a list with 5000 elements
trial <- lapply(trial, function (x) rpois(3,42)) # Fill each list element with 3 random draws
means <- sapply(trial, mean) # Take the mean for each (our MLE)
qplot(means, geom = "histogram") # Show
```

Play around with the number of observations and Poisson parameter.

```{r}

nb <- rnbinom(1000, size = 1, mu = 5) # In R, as size -> infinity, NB ~ Pois
p <- rpois(1000, 5)
qplot(nb)
qplot(p)

```


```{r}
fitdistr(poisson.values, "negative binomial")
```

